Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-02-26 parallel-tractability-J-v9/conclusions.tex
finished checking the appendix, merges changes into rel. work and conclusions

In this paper, we studied the problem of parallel tractability of
materialization on the datalog rewritable ontologies.
To identify the parallel tractability classes,
we proposed two \texttt{NC} algorithms, Algorithm~$\mathsf{A}_{\text{bsc}}^\psi$
and Algorithm~$\mathsf{A}_{\text{opt}}^\psi$, that perform materialization on
datalog rewritable ontology languages.
Based on these algorithms, we identified the corresponding
parallel tractability classes such that materialization
on the datalog programs in these classes is in \texttt{NC} complexity.
We further studied two specific ontology languages, DL-Lite and DHL (including one of its extension).
We showed that any ontology expressed in DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ is tractable in parallel.
For DHL and DHL($\circ$), we proposed two restrictions such that materialization is tractable in parallel.

We analyzed the benchmark LUBM and the real-world dataset YAGO
and gave the cases where these ontologies belong to parallel tractability classes.
On the other hand, we used an optimization strategy based on SWD paths
to give a practical algorithm variant Algorithm~$\mathsf{A}_{\text{prc}}$, which can
also be restricted to an \texttt{NC} version.
We implemented a system based on Algorithm~$\mathsf{A}_{\text{prc}}$ and compared it to
the state-of-the-art reasoner RDFox.
The experimental results showed that the optimizations proposed in this paper resulted in a
better performance on ontologies that were
tractable in parallel compared with RDFox.

In the future work, we will extend our work in two lines.
One line is a further study of the parallel tractability of
other expressive ontology languages, in addition to
datalog rewritable ontology languages. Since different expressive OWL languages
have different syntaxes and higher reasoning complexities than \texttt{PTime}-complete, we need to
explore more restrictions that are practical and make materialization
tractable in parallel. Another line of the future work is to
study how to further apply the
results in practice.
One idea is to apply the technique of SWD paths to enhance other reasoning-based tasks,
like ontology classification, ontology debugging and query answering.

>>>>

In this paper, we studied the problem of parallel tractability of
materialization on datalog rewritable ontologies.
To identify the parallel tractability classes,
we proposed two \texttt{NC} algorithms, Algorithm~$\mathsf{A}_{\text{bsc}}^\psi$
and Algorithm~$\mathsf{A}_{\text{opt}}^\psi$, that perform materialization on
datalog rewritable ontology languages.
Based on these algorithms, we identified the corresponding
parallel tractability classes such that materialization
on the datalog programs in these classes is in \texttt{NC} complexity.
We further studied two specific ontology languages, DL-Lite and DHL (including one of its extension).
We showed that any ontology expressed in DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ is tractable in parallel.
For DHL and DHL($\circ$), we proposed two restrictions such that materialization is tractable in parallel.

We analyzed the benchmark LUBM and the real-world dataset YAGO
and gave the cases where these ontologies belong to parallel tractability classes.
On the other hand, we used an optimization strategy based on SWD paths
to give a practical algorithm variant Algorithm~$\mathsf{A}_{\text{prc}}$, which can
also be restricted to an \texttt{NC} version.
We implemented a system based on Algorithm~$\mathsf{A}_{\text{prc}}$ and compared it to
the state-of-the-art reasoner RDFox.
The experimental results showed that the optimizations proposed in this paper result in a
better performance on ontologies that are
tractable in parallel compared to RDFox.

In future work, we plan to extend our work in two lines.
One line is a further study of parallel tractability of
other expressive ontology languages, in addition to
datalog rewritable ontology languages. Since different expressive OWL languages
have different syntaxes and higher reasoning complexities than \texttt{PTime}-complete, we need to
explore more restrictions that are practical and make materialization
tractable in parallel. Another line of work is to
study how to further apply the
results in practice.
One idea is to apply the technique of SWD paths to enhance other
reasoning-based tasks such as ontology classification, ontology
debugging and query answering.

### 2018-02-26 parallel-tractability-J-v9/optimizedAlgo.tex
finished checking the appendix, merges changes into rel. work and conclusions

\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
\item add \texttt{rch}$(A(a),B(a))$ to $S_{\textit{\!\tiny rch}}$ where $A(a)\rightarrow B(a)\in P^*$, $\mathcal{O}\models A\sqsubseteq_* B$ and $A(a)\in V$;

>>>>

\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
\item add \texttt{rch}$(A(a),B(a))$ to $S_{\textit{\!\tiny rch}}$
  where $A(a)\rightarrow B(a)\in P^*$, $\mathcal{O}\models
  A\sqsubseteq_* B$ \BG{By referring to the entailment relation, the *
  seems not needed: $\mathcal{O}\models
  A\sqsubseteq B$. Similarly below. I noticed that, when checking
  whether we introduced the * notation with a clear reference to the
  ontology/RBox, but we no not. I added a sentence to say that we omit
the RBox (as we never introduced the * w.r.t.\ the whole ontology) if
it is clear from the context. } and $A(a)\in V$;

### 2018-02-26 parallel-tractability-J-v9/parallelTractabilityOntMat.tex
Proof-read eval section
finished checking the appendix, merges changes into rel. work and conclusions

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{fig-pathtwist.pdf}
\caption{The materialization graph of $\mathcal{O}_{\text{ex}_6}$
  (left) and of $\mathcal{O}_{\text{ex}_7}$ (right)}
\label{fig:ex3_4}
\end{center}
\end{figure}

@@

\begin{definition}\label{def:dhlplus}
$\mathcal{D}_{\textit{\text{dhl}}(\circ)}$ is a class of datalog programs where each program
is rewritten from a DHL($\circ$) ontology and the following
conditions are satisfied:
\begin{enumerate}[leftmargin=4ex,label=\arabic*.]
\item for all axioms of the form $A_1\sqcap A_2\sqsubseteq B$,
    at least one concept of $A_1$ and $A_2$ is a \emph{simple concept};
\item for all axioms of the form $R_1\circ R_2\sqsubseteq R$,
    if there does exits a transitive role $S$ such that
    $R\sqsubseteq_* S$, then both $R_1$ and $R_2$ are \emph{simple
      roles}; otherwise at least one of $R_1$ and $R_2$ is a \emph{simple role}.
\end{enumerate}
\end{definition}

@@

In this part, we discuss parallel tractability of reasoning over RDFS ontologies.
Although RDFS is not directly based on a description logic, it can (partly) be described by a set of DL axioms \cite{GrosofHVD03}.
The correspondence between RDFS statements and DL axioms is listed in Table~\ref{tab:rdfs}.
We call RDFS statements of the form (1--4) \emph{the schema data}
(which correspond to TBox axioms), statements of the form~(5) and~(6)
\emph{the instance data} (which correspond to ABox axioms).

\begin{table}
\centering
\caption{RDFS statements and the corresponding DL axioms}
{\setlength{\tabcolsep}{6mm}
\begin{tabular}{lcc}
\hline
& RDFS statements & Axiom\\
\hline
\hline
(1)&$P~~\texttt{rdfs:domain}~~C$& $\top\sqsubseteq\forall P^-.C$\\

(2)&$P~~\texttt{rdfs:range}~~C$& $\top\sqsubseteq\forall P.C$\\

(3)&$B~~\texttt{rdfs:subClassOf}~~C$& $B\sqsubseteq C$\\

(4)&$R~~\texttt{rdfs:subPropertyOf}~~S$& $R\sqsubseteq S$\\
\hline
(5)&$a~~\texttt{rdf:type}~~C$& $C(a)$\\

(6)&$a~~R~~b$& $R(a,b)$\\
\hline
\end{tabular}}
\label{tab:rdfs}
\end{table}

>>>>

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{fig-pathtwist.pdf}
\caption{The materialization graph of $\mathcal{O}_{\text{ex}_6}$
  (left) and of $\mathcal{O}_{\text{ex}_7}$ (right)}
\label{fig:ex3_4}
\end{center}
\end{figure}

@@

\begin{definition}\label{def:dhlplus}
$\mathcal{D}_{\textit{\text{dhl}}(\circ)}$ is a class of datalog programs where each program
is rewritten from a DHL($\circ$) ontology and the following
conditions are satisfied:
\begin{enumerate}[leftmargin=4ex,label=\arabic*.]
\item for all axioms of the form $A_1\sqcap A_2\sqsubseteq B$,
    at least one concept of $A_1$ and $A_2$ is a \emph{simple concept};
\item for all axioms of the form $R_1\circ R_2\sqsubseteq R$,
    if there exits a transitive role $S$ such that
    $R\sqsubseteq_* S$, then both $R_1$ and $R_2$ are \emph{simple
      roles}; otherwise at least one of $R_1$ and $R_2$ is a \emph{simple role}.
\end{enumerate}
\end{definition}

@@

In this part, we discuss parallel tractability of reasoning over RDFS ontologies.
Although RDFS is not directly based on a description logic, it can (partly) be described by a set of DL axioms \cite{GrosofHVD03}.
The correspondence between RDFS statements and DL axioms is listed in Table~\ref{tab:rdfs}.
We call RDFS statements of the form (S1--S4) \emph{the schema data}
(which correspond to TBox axioms), statements of the form~(S5) and~(S6)
\emph{the instance data} (which correspond to ABox axioms).

\begin{table}
\centering
\caption{RDFS statements and the corresponding DL axioms}
{\setlength{\tabcolsep}{6mm}
\begin{tabular}{lcc}
\hline
& RDFS statements & Axiom\\
\hline
\hline
(S1)&$P~~\texttt{rdfs:domain}~~C$& $\top\sqsubseteq\forall P^-.C$\\

(S2)&$P~~\texttt{rdfs:range}~~C$& $\top\sqsubseteq\forall P.C$\\

(S3)&$B~~\texttt{rdfs:subClassOf}~~C$& $B\sqsubseteq C$\\

(S4)&$R~~\texttt{rdfs:subPropertyOf}~~S$& $R\sqsubseteq S$\\
\hline
(S5)&$a~~\texttt{rdf:type}~~C$& $C(a)$\\

(S6)&$a~~R~~b$& $R(a,b)$\\
\hline
\end{tabular}}
\label{tab:rdfs}
\end{table}

### 2018-02-26 parallel-tractability-J-v9/relatedWork.tex
finished checking the appendix, merges changes into rel. work and conclusions

The parallel reasoner RDFox \cite{MotikNPHO14} is used to evaluate our implementation. RDFox is a state-of-the-art system that handles reasoning on datalog rewritable ontology languages. Algorithm~$\mathsf{A}_{bsc}$ proposed in Section~\ref{sec:ptclass}, is similar to the main algorithm of RDFox (see \cite{MotikNPHO14}, Sections~3 and 4). The main difference is that a group of rule instantiations is handled by one processor (namely a thread) in RDFox, while in Algorithm~$\mathsf{A}_{bsc}$, each rule instantiation is assigned to a unique processor. Thus, in theory, the materialization of the datalog program in Example~\ref{exp:mg} is serial in RDFox {\color{red}The example has two rules. Why can they not be parallelized in RDFox?}. ItThe experiments also show that parallelism does not lead to a remarkable improvement for RDFox for the ontology series not in $\mathcal{D}_{\textit{\text{dhl}}(\circ)}$. We use Algorithm~$\mathsf{A}_{opt}$ to show that the ontology in Example~\ref{exp:mg} is also tractable in parallel. Our implementation ParallelDHL, which is based on Algorithm~$\mathsf{A}_{prc}$, also shows a better performance than RDFox when handling the test ontologies that belong to $\mathcal{D}_{\textit{\text{dhl}}(\circ)}$.

>>>>

The parallel reasoner RDFox \cite{MotikNPHO14} is also used in the evaluation of our implementation. RDFox is a highly optimized system for reasoning over datalog rewritable ontology languages. Algorithm~$\mathsf{A}_{\text{bsc}}$ proposed in Section~\ref{sec:ptclass} is similar to the main algorithm of RDFox (see \cite{MotikNPHO14}, Sections~3 and 4). The main difference is that a group of rule instantiations is handled by one processor (namely a thread) in RDFox, while in Algorithm~$\mathsf{A}_{\text{bsc}}$, each rule instantiation is assigned to a unique processor.
Analogous to the method used in RDFox, ParallelDHL is also implemented by assigning a group of
rule instantiations to a processor.
Specially, ParallelDHL is also optimized by using SWD paths for handling
ontologies that belong to parallel tractability classes.
The experimental results show that this optimization results in a
better performance on ontologies that are
tractable in parallel compared with RDFox.
